---
title: "SOCS0100-ASSIGNMENT"
format: html
editor: visual
---

# Introduction

Nowadays, urban crime is a persistent social issue which affects the society's public safety, inequality, residents' everyday experiences and so on. As one of Europe's largest and most diverse cities, London exhibits significant and concerning crime issues. Understanding how crime is distributed across space and time is essential for not only informing evidence-based policy and analysis to address the problem but also for improving residents' sense of security and will-being.

Therefore, this project examines crime patterns in London, particularly in 2024, incorporating police station locations to provide additional contextual understanding. It first presents the data collection process of two datasets, followed by data wrangling and the development of a Shiny dashboard for crime visualisations. Finally, it reflects critically on the use of Copilot.

```{r}
#| include: false

#Setup
if (!require("pacman")) {
  install.packages("pacman")
}
pacman::p_load(httr, jsonlite, dplyr, ggplot2, purrr, sf, lubridate, tidyr, osmdata)


library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(purrr)
library(sf)
library(lubridate)
library(tidyr)
library(osmdata)


```

# Part I-A Automated Data Collection

To explore crime, this project collects two datasets: street-level crime records from the UK Police API and police station location data from OpenStreetMap (OSM). API technique is selected to do the automated data collection. Output sections for the two collected datasets show the variables contained in the datasets and the corresponding values of the first six observations, to illustrate stuctures of the datasets.

Firstly, since the crime data contains longitude and latitude, a spatial bounding box based on approximate London coordinates was defined to collect crime data for the London area. This supports API querying, as the UK Police Street-level Crime API requires a polygon (poly) parameter defined by longitude and latitude. A grid-based polygon approach was used to keep polygons small, reducing the risk of exceeding API limits, although this increases request time.

```{r}
#| eval: false

#Create London polygon for data collection with longitude and latitude


lat_min <- 51.38
lat_max <- 51.67
lng_min <- -0.56
lng_max <- 0.28

make_grid_polys <- function(lat_min, lat_max, lng_min, lng_max, step = 0.05) {
  
  lats <- seq(lat_min, lat_max, by = step)
  lngs <- seq(lng_min, lng_max, by = step)
  
  polys <- list()
  
  for (i in seq_len(length(lats) - 1)) {
    for (j in seq_len(length(lngs) - 1)) {
      
      poly <- paste(
        paste(lats[i],   lngs[j],   sep = ","),
        paste(lats[i],   lngs[j+1], sep = ","),
        paste(lats[i+1], lngs[j+1], sep = ","),
        paste(lats[i+1], lngs[j],   sep = ","),
        sep = ":"
      )
      
      polys <- append(polys, poly)
    }
  }
  
  polys
}

london_polys <- make_grid_polys(
  lat_min = 51.38,
  lat_max = 51.67,
  lng_min = -0.56,
  lng_max = 0.28,
  step = 0.07
)
```

To request data with the UK Police API, a safe_get_json function is implemented to handle potential failures, errors, or empty responses when retrieving large volumes of data. Instead of interrupting execution, the function allows the collection process to continue, ensuring a smoother multi-month, multi-polygon data collection process.

A second function, get_crime_poly, ensures API compliance by requiring both polygon and date parameters. After defining the date (all months from 2020-2024), the data is collected iteratively by looping over all London grid polygons and monthly dates. The messages help monitor the collection process and identify problematic polygon or date if errors occur. Lastly, the poly and date variables are appended to each record to facilitate subsequent analysis.

Finally, the crime dataset is constructed by binding the results across all polygon–month combinations.

::: panel-tabset
## Code 1

```{r}
#| eval: false

#Collect London crime data

#Use a function to safely request JSON data from the API, and avoid crashing when errors or empty responses return
safe_get_json <- function(url) {
  res <- GET(url)
  status <- status_code(res)
  txt <- content(res, "text", encoding = "UTF-8")
  
  if (status != 200) {
    message("Status ", status, " for URL: ", url)
    return(data.frame())
  }
  
  if (grepl("^\\s*<", txt)) {
    message("HTML returned for URL: ", url)
    return(data.frame())
  }
  
  if (nchar(txt) < 2) {
    message("Empty response for URL: ", url)
    return(data.frame())
  }
  
  tryCatch(
    fromJSON(txt),
    error = function(e) {
      message("JSON parse error for URL: ", url)
      return(data.frame())
    }
  )
}

#Build the API request for a given polygon and month, then use the previous function to collect crime data
get_crime_poly <- function(poly, date) {
  url <- paste0(
    "https://data.police.uk/api/crimes-street/all-crime?",
    "poly=", URLencode(poly),
    "&date=", date
  )
  safe_get_json(url)
}

#Create a sequence of months needed
dates <- seq(as.Date("2020-01-01"), as.Date("2024-12-01"), by = "month")
dates <- format(dates, "%Y-%m")


#Loop through every polygon and month, collect the data and store it
all_results <- list()
counter <- 1

for (p in london_polys) {
  for (d in dates) {
    
    message("Fetching poly ", counter, "/", length(london_polys), " for date ", d)
    
    df <- get_crime_poly(p, d)
    
    if (nrow(df) > 0) {
      df$poly_id <- counter
      df$date <- d
      all_results[[length(all_results) + 1]] <- df
    }
    
    Sys.sleep(0.5)
  }
  
  counter <- counter + 1
}

crime_data <- bind_rows(all_results)

write.csv(crime_data, "data/london_crime_data.csv", row.names = FALSE)


```

## Output 1

```{r}
#| echo: false

crime_data <- read.csv("data/london_crime_data.csv")

head(crime_data)
```
:::

To complement the crime dataset with information on the spatial distribution of policing infrastructure, the wrapper provided by osmdata package is used to collect data of police station location. The collected data includes point data and polygon data, and they are saved respectively.

```{r}
#| include: false

police_point_data <- st_read("data/london_police_points.geojson")
police_polygon_data <- st_read("data/london_police_polygons.geojson")
```

::: panel-tabset
## Code 2

```{r}
#| eval: false

#Collect data of police stations' location

#Use the wrapper from r package
bb <- getbb ("london uk")
london_rst <- opq(bb)%>%
  add_osm_feature(key = "amenity", value = "police") %>%
  osmdata_sf() 

london_police_points <- london_rst$osm_points %>% 
  select(name, geometry)

london_police_polygons <- london_rst$osm_polygons %>%
  select(name, geometry)


st_write(london_police_points, "data/london_police_points.geojson")

st_write(london_police_polygons, "data/london_police_polygons.geojson")

```

## Output 2

```{r}
#| echo: false


head(police_point_data)
head(police_polygon_data)
```
:::

```{r}
#| include: false

crime_data <- read.csv("data/london_crime_data.csv")

```

# Part I-B Data Exploration and Contextualisation

After data collection, the dataset is cleaned and processed. Three datasets and one function are created from the collected data to support plotting and mapping in the Shiny app. Again, output sections display the variables contained in the datasets and the corresponding values of the first six observations, to illustrate stuctures of the datasets.

Firstly, the raw dataset is filtered to retain only crime recorded in 2024 for selected crime categories. Then the variables month, longitude, latitude, and street name are cleaned to make the analysis of crime patterns more easier. Lastly, key variables are selected to create the ranking_data for later uses. To clarify, street_clean variable indicates the street or area where the crime occurred or nearby. This dataset will be used for a function to rank streets or areas. It contains 630725 observations and 6 variables.

::: panel-tabset
## Code 1

```{r}
#| eval: false

#Define the crime categories focused for this project
crime_categories <- c(
  "all-crime",
  "anti-social-behaviour",
  "bicycle-theft",
  "criminal-damage-arson",
  "drugs",
  "possession-of-weapons",
  "public-order",
  "robbery",
  "violent-crime"
)


crime_2024 <- crime_data %>%
  filter(
    category %in% crime_categories,
    substr(date, 1, 4) == "2024"
  ) %>%
  mutate(
    month = factor(substr(date, 6, 7), levels = sprintf("%02d", 1:12))
  )

crime_2024_clean <- crime_2024  %>%
  rename(
    latitude = location.latitude,
    longitude = location.longitude
  )

ranking_data <- crime_2024_clean %>%
  mutate(
    street_clean = gsub("On or near ", "", location.street.name, perl = TRUE)
  ) %>%
  select(
    category,
    latitude,
    longitude,
    month,
    date,
    street_clean
  )

write.csv(ranking_data, "data/ranking_data.csv", row.names = FALSE)
```

## Output 1

```{r}
#| echo: false

crime_categories <- c(
  "all-crime",
  "anti-social-behaviour",
  "bicycle-theft",
  "criminal-damage-arson",
  "drugs",
  "possession-of-weapons",
  "public-order",
  "robbery",
  "violent-crime"
)


crime_2024 <- crime_data %>%
  filter(
    category %in% crime_categories,
    substr(date, 1, 4) == "2024"
  ) %>%
  mutate(
    month = factor(substr(date, 6, 7), levels = sprintf("%02d", 1:12))
  )

crime_2024_clean <- crime_2024  %>%
  rename(
    latitude = location.latitude,
    longitude = location.longitude
  )

ranking_data <- crime_2024_clean %>%
  mutate(
    street_clean = gsub("On or near ", "", location.street.name, perl = TRUE)
  ) %>%
  select(
    category,
    latitude,
    longitude,
    month,
    date,
    street_clean
  )

write.csv(ranking_data, "data/ranking_data.csv", row.names = FALSE)

head(ranking_data)
```
:::

This function generates street rank data for Shiny app. It filters crime category and counts of crime on each street, it then rank the street according to number of a particular crime category in 2024. This allows identification of local crime hotspots and facilitates comparative analysis across different types of crime.

```{r}
#| eval: false

get_street_ranking <- function(crime_type,
                               month = "all",
                               top_n = 10) {
  
  df <- ranking_data
  
  df <- df %>% filter(category == crime_type)
  
  df %>%
    count(street_clean, sort = TRUE) %>%
    slice_head(n = top_n)
}
```

For trend data, the monthly crime frequency for each crime category is computed by grouping category and month and summarising the total count of incidents within each group from the 2024 dataset. This will be used for comparison and trend plots in Shiny app. It contains 96 observations and 3 variables.

::: panel-tabset
## Code 2

```{r}
#| eval: false

trend_data <- crime_2024 %>%
  group_by(category, month) %>%
  summarise(count = n(), .groups = "drop")

write.csv(trend_data, "data/trend_data.csv", row.names = FALSE)

```

## Output 2

```{r}
#| echo: false

crime_2024 <- crime_data %>%
  filter(
    category %in% crime_categories,
    substr(date, 1, 4) == "2024"
  ) %>%
  mutate(
    month = factor(substr(date, 6, 7), levels = sprintf("%02d", 1:12))
  )

trend_data <- crime_2024 %>%
  group_by(category, month) %>%
  summarise(count = n(), .groups = "drop")

write.csv(trend_data, "data/trend_data.csv", row.names = FALSE)

head(trend_data)
```
:::

Lastly, a mapping dataset is constructed by selecting coordinates, crime category, and month variables from the crime_2024_clean dataset. This will be used to map the data on an interactive map in Shiny app, which contains 630725 observations and 4 variables.

::: panel-tabset
## Code 3

```{r}
#| eval: false

map_data <- crime_2024_clean %>%
  select(longitude, latitude, category, month)

write.csv(map_data, "data/map_data.csv", row.names = FALSE)

```

## Output 3

```{r}
#| echo: false

map_data <- crime_2024_clean %>%
  select(longitude, latitude, category, month)

write.csv(map_data, "data/map_data.csv", row.names = FALSE)

head(map_data)
```
:::

Overall, these datasets have substantial social science significance. Visualisations produced using these datasets enable spatial analysis of crime, spatial inequality and so on. The inclusion of police station locations further supports examination of policing resources, accessibility, and their relationship to crime patterns. Together, the datasets provide a rich empirical foundation for policy-making aimed at improving socio-economic outcomes. Additionally, individuals can use the maps and plots to avoid higher-risk areas and improve personal safety.

However, these datasets have limitations. In the data collection process, despite dividing London into many small polygons, the Police API will return a 503 error if a custom area contains over 10,000 crimes, which leads to missing records in some polygons, such as around Brixton. Similarly, the police station data from OSM are incomplete. These non-random missingness may affect the accuracy and significance of subsequent analyses and visualisations.

# Part II-A Building an Interactive Dashboard with R Shiny

Using the organised datasets, an interactive Shiny dashboard is produced. It contains three complementary analytical sections, each designed to show different dimensions of crime patterns in London in 2024. Although the dataset covers records from 2020 to 2024, given the large amount of information about crime and my intention to be more focused, this Shiny app only explore 2024 crime patterns.

The crime map visualises the spatial distribution of crime by category using map_data with clustered point markers due to the large amount of points. This enables users to identify spatial hotspots and observe how crime clusters vary across places. The optional overlay of police station locations provides additional spatial contexts which enables users to compare crime pattern and policing infrastructure. Based on the map, Central London and East London exhibit higher crime counts than other parts. While it is also worth noting that the highest concentration of police stations is also located in Central London.

The top streets section identifies areas or streets with the highest crime counts using ranking_data. Bar charts are used to rank as they enable clear comparisons. The results reveal that parking areas record the highest number of crime incidents, followed by supermarkets and shopping areas. An interactive pie chart further illustates crime categories within a selected high-crime area. Across the top ten areas, violent-crime consistently represents the largest proprotion (\>34% for all), followed by anti-social behaviour.

The last section, the temporal patterns section, explores how crime varies over time using trend_data. Line charts are used to display monthly trends for specific crime categories. It shows that there are seasonal variations for all crime types. The monthly bar chart enables cross-category comparison, illustating how the distribution of crime types varies over time.

Overall, the dashboard integrates spatial, categorical, and temporal perspectives, using interactive visualisations to improve understanding of crime patterns in London and urban crime data.

# Part II-C Critical Engagement with AI

Throughout this project, Copilot is used as a collaborative tool to interpret error messages, clarify computational logic, and refine codes. It provides valuable suggestions to make the data-collection process smoother and more convenient, and it also supported to make the Shiny app more interactive and visually appealing.

For example, Copilot supports to write a safer safe_get_json wrapper so that API errors will not break the whole loop. Before this function was applied, the data collection process encountered multiple failures caused by unstable API responses during large-scale requests.

```{r}
#| eval: false

safe_get_json <- function(url) {
  res <- GET(url)
  status <- status_code(res)
  txt <- content(res, "text", encoding = "UTF-8")
  
  if (status != 200) {
    message("Status ", status, " for URL: ", url)
    return(data.frame())
  }
  
  if (grepl("^\\s*<", txt)) {
    message("HTML returned for URL: ", url)
    return(data.frame())
  }
  
  if (nchar(txt) < 2) {
    message("Empty response for URL: ", url)
    return(data.frame())
  }
  
  tryCatch(
    fromJSON(txt),
    error = function(e) {
      message("JSON parse error for URL: ", url)
      return(data.frame())
    }
  )
}
```

Also, progress messages like "Fetching poly" are suggested to keep track of long runs since the whole collection process took more than 1 hour.

```{r}
#| eval: false

message("Fetching poly ", counter, "/", length(london_polys), " for date ", d)
```

However, although the function allowed the data‑collection process to continue despite failures and printed messages to indicate them, there were so many appearing quickly in the console that checking and fixing each was impossible. Consequently, the final dataset inevitably includes many unsuccessful requests that are non-random missingness which couldn’t be properly reviewed.

When making map in Shiny app using all the crime records collected, the initial map was completely overcrowded with individual points and was impossible to see spatial pattern. Copilot suggested using marker clustering, which makes the distribution of crime clearer to see.

```{r}
#| eval: false

clusterOptions = markerClusterOptions()
```

Moreover, the first Shiny app version used a simple titlePanel function, while a more flexible fluidRow function is recommended for a more polished interface. Copilot also guided how to add an imported image to the app.

Before:

```{r}
#| eval: false

 # App title
  titlePanel("London Crime Dashboard (2024)"),
  
  # Intro text
  h3("Welcome to the London Crime Dashboard"),
  "This dashboard provides a comprehensive overview of crime patterns across London in 2024.",
```

After:

```{r}
#| eval: false

 # App title
  fluidRow(
    column(
      width = 12,
      h1("London Crime Dashboard (2024)",
         style = "margin-top: 20px; font-weight: 700;")
    )
  ),
  

  fluidRow(
    column(
      width = 4,
      img(
        src = "image.jpg", 
        width = "100%",
        style = "border-radius: 10px; margin-top: 10px;"
      )
    ),
    column(
      width = 8,
      h3("Welcome to the London Crime Dashboard"),
      p("This dashboard provides a comprehensive overview of crime patterns across London in 2024."),
```

Lastly, Copilot also supported with improving interactivity. By using renderPlotly function, the corresponding crime count is displayed when users hover over a bar, which makes the static plots more interactive and detailed.

Before:

```{r}
#| eval: false

output$bar_plot <- renderPlot({
    
    df <- trend_data %>%
      mutate(month = sprintf("%02d", as.integer(as.character(month)))) %>%
      filter(month == input$month)
    
    ggplot(df, aes(x = reorder(category, count), y = count, fill = category)) +
      geom_col() +
      coord_flip() +
      labs(
        title = paste("Crime Category Comparison in Month", input$month),
        x = "Crime Category",
        y = "Crime Count"
      )
})
```

After:

```{r}
#| eval: false

output$bar_plot <- renderPlotly({
    
    df <- trend_data %>%
      mutate(month = sprintf("%02d", as.integer(as.character(month)))) %>%
      filter(month == input$month)
    
    p <- ggplot(df, aes(
      y = reorder(category, count),
      x = count,
      fill = category,
      text = paste("<br>Count:", count)
    )) +
      geom_col() +
      theme_minimal() +
      theme(legend.position = "none") +
      labs(
        title = paste("Crime Category Comparison in Month", input$month),
        x = "Crime Count",
        y = "Crime Category"
      )
    
    ggplotly(p, tooltip = "text") %>%
      config(displayModeBar = FALSE)
  })
```

Overall, using Copilot improved my code proficiency and enabled a deeper understanding of automated data collection and Shiny app building process by providing many insightful perspectives and refined codes.
